---
title: "2 Data Import and Wrangling"
author: "Mohan Khanal"
date: "2026-01-08"
format:
  html:
    toc: true          # Enable table of contents
    toc-location: left # Place TOC on the left side
    toc-title: "Contents"
    toc-depth: 2       # Show headers up to level 2 (##)
  pdf:
    toc: true          # Also enable TOC in PDF
    number-sections: true
---

# 📘 Chapter 2: Data Import and Data Wrangling

## 🕐 Agenda

-   **Part 1 (30 mins)** – Data Import and Data Wrangling (Reading Different Types of Data)
-   **Part 2 (40 mins)** – Data Cleaning and Manipulation using `dplyr`
-   **Part 3 (30 mins)** – Handling Missing Data
-   **Part 4 (20 mins)** – Practice and Q&A

------------------------------------------------------------------------

## Part 1: Reading Different Types of Data (30 mins)

We can also check our working directory using the command getwd().

### CSV File Example: Iris Dataset

```{r}
# We can import csv file using the base library using the command read.csv 
iris_data <- read.csv("../data/Iris.csv")

# Or we can use readr library using the command read_csv to import csv 
library(readr)
iris_data_r <- read_csv("../data/Iris.csv")

# View column names
names(iris_data)

# View first few rows
head(iris_data)

# Check structure
str(iris_data)
```

### Understanding the Difference: read.csv() vs read_csv()

```{r}
# Base R: read.csv()
class(iris_data)        # Returns "data.frame"

# readr: read_csv()
class(iris_data_r)      # Returns "spec_tbl_df" "tbl_df" "tbl" "data.frame"

# Both work, but read_csv() is generally faster and has better defaults
```

### Quick Data Exploration

```{r}
# Dimensions
dim(iris_data)

# Summary statistics
summary(iris_data)

# View in RStudio
# View(iris_data)
```

------------------------------------------------------------------------

## Part 2: Data Cleaning and Manipulation using `dplyr` (40 mins)

### Load Required Packages

```{r, warning=FALSE}
library(dplyr)
```

### The Pipe Operator: %>%

The pipe operator `%>%` makes code more readable by chaining operations together.

```{r}
# Without pipe (hard to read)
head(filter(iris_data, SepalLengthCm > 6))

# With pipe (easier to read)
iris_data %>%
  filter(SepalLengthCm > 6) %>%
  head()
```

**Keyboard shortcut**: Ctrl + Shift + M (Windows) or Cmd + Shift + M (Mac)

### Basic Wrangling Examples

#### Selecting Columns with `select()`

```{r, warning=FALSE}
# Select specific columns
iris_data %>%
  select(Species, SepalLengthCm, PetalLengthCm) %>%
  head()

# Select columns using patterns
iris_data %>%
  select(starts_with("Petal")) %>%
  head()

iris_data %>%
  select(ends_with("Cm")) %>%
  head()

# Exclude columns with minus sign
iris_data %>%
  select(-Id) %>%
  head()
```

#### Filtering Rows with `filter()`

```{r, warning=FALSE}
# Filter for setosa species only
iris_data %>%
  filter(Species == "Iris-setosa") %>%
  head()

# Filter flowers with long petals
iris_data %>%
  filter(PetalLengthCm > 5) %>%
  head()

# Multiple conditions with AND (&)
iris_data %>%
  filter(Species == "Iris-virginica" & SepalLengthCm > 6.5) %>%
  head()

# Multiple conditions with OR (|)
iris_data %>%
  filter(PetalLengthCm > 6 | PetalWidthCm > 2) %>%
  head()

# Using %in% operator for multiple values
iris_data %>%
  filter(Species %in% c("Iris-setosa", "Iris-versicolor")) %>%
  head()
```

#### Creating New Variables with `mutate()`

```{r, warning=FALSE}
# Calculate areas
iris_data <- iris_data %>%
  mutate(
    SepalArea = SepalLengthCm * SepalWidthCm,
    PetalArea = PetalLengthCm * PetalWidthCm
  )

head(iris_data)

# Create categorical variables
iris_data <- iris_data %>%
  mutate(
    SizeCategory = ifelse(PetalLengthCm > 4, "Large", "Small"),
    SepalRatio = SepalLengthCm / SepalWidthCm
  )

head(iris_data)

# Using case_when() for multiple conditions
iris_data <- iris_data %>%
  mutate(
    PetalSize = case_when(
      PetalLengthCm < 2 ~ "Small",
      PetalLengthCm < 5 ~ "Medium",
      TRUE ~ "Large"
    )
  )

head(iris_data)
```

#### Sorting Rows with `arrange()`

```{r, warning=FALSE}
# Sort by Petal Length (ascending)
iris_data %>%
  arrange(PetalLengthCm) %>%
  head()

# Sort by Petal Length (descending)
iris_data %>%
  arrange(desc(PetalLengthCm)) %>%
  head()

# Sort by multiple columns
iris_data %>%
  arrange(Species, desc(SepalLengthCm)) %>%
  head()
```

#### Summarizing with `group_by()` and `summarize()`

```{r, warning=FALSE}
# Overall summary
iris_data %>%
  summarize(
    avg_sepal_length = mean(SepalLengthCm, na.rm = TRUE),
    avg_petal_length = mean(PetalLengthCm, na.rm = TRUE),
    total_flowers = n()
  )

# Summary by species
iris_data %>%
  group_by(Species) %>%
  summarize(
    avg_sepal_length = mean(SepalLengthCm, na.rm = TRUE),
    avg_sepal_width = mean(SepalWidthCm, na.rm = TRUE),
    avg_petal_length = mean(PetalLengthCm, na.rm = TRUE),
    avg_petal_width = mean(PetalWidthCm, na.rm = TRUE),
    count = n()
  )

# Summary by multiple groups
iris_data %>%
  group_by(Species, SizeCategory) %>%
  summarize(
    avg_sepal_length = mean(SepalLengthCm, na.rm = TRUE),
    count = n(),
    .groups = 'drop'
  )
```

### Combining Multiple Operations

```{r, warning=FALSE}
# Complex pipeline: Find large virginica flowers and their stats
iris_data %>%
  filter(Species == "Iris-virginica", PetalLengthCm > 6) %>%
  select(Species, SepalLengthCm, PetalLengthCm, SepalArea, PetalArea) %>%
  arrange(desc(PetalArea)) %>%
  head(10)
```

### Additional Useful Functions

#### Count rows with `count()`

```{r}
# Count flowers by species
iris_data %>%
  count(Species)

# Count with sorting
iris_data %>%
  count(Species, sort = TRUE)

# Count by multiple variables
iris_data %>%
  count(Species, SizeCategory)
```

#### Get unique values with `distinct()`

```{r}
# Get unique species
iris_data %>%
  distinct(Species)

# Get unique combinations
iris_data %>%
  distinct(Species, PetalSize)
```

------------------------------------------------------------------------

## Part 3: Handling Missing Data (30 mins)

### Checking Missing Values

```{r}
# Load naniar for missing data visualization
library(naniar)

# Check for missing values
miss_var_summary(iris_data)

# Check if any values are missing
any(is.na(iris_data))

# Count NAs in each column
colSums(is.na(iris_data))
```

### Creating Sample Data with Missing Values

For demonstration, let's introduce some missing values:

```{r}
# Create a copy with missing values
iris_missing <- iris_data
set.seed(123)  # For reproducibility

# Randomly introduce NAs
iris_missing$SepalLengthCm[sample(1:nrow(iris_missing), 10)] <- NA
iris_missing$PetalWidthCm[sample(1:nrow(iris_missing), 15)] <- NA

# Check missing data
miss_var_summary(iris_missing)
```

### Removing Missing Values

```{r}
# Remove rows with any missing values
iris_complete <- na.omit(iris_missing)

cat("Original rows:", nrow(iris_missing), "\n")
cat("After removing NAs:", nrow(iris_complete), "\n")

# Remove only if specific column has NA
iris_filtered <- iris_missing %>%
  filter(!is.na(SepalLengthCm))

cat("After filtering SepalLengthCm:", nrow(iris_filtered), "\n")
```

### Replacing Missing Values

```{r}
# Replace NAs with mean
iris_mean <- iris_missing %>%
  mutate(
    SepalLengthCm = ifelse(is.na(SepalLengthCm), 
                          mean(SepalLengthCm, na.rm = TRUE), 
                          SepalLengthCm),
    PetalWidthCm = ifelse(is.na(PetalWidthCm), 
                         mean(PetalWidthCm, na.rm = TRUE), 
                         PetalWidthCm)
  )

# Verify
miss_var_summary(iris_mean)

# Replace with median (more robust to outliers)
iris_median <- iris_missing %>%
  mutate(
    SepalLengthCm = ifelse(is.na(SepalLengthCm), 
                          median(SepalLengthCm, na.rm = TRUE), 
                          SepalLengthCm),
    PetalWidthCm = ifelse(is.na(PetalWidthCm), 
                         median(PetalWidthCm, na.rm = TRUE), 
                         PetalWidthCm)
  )

# Replace with species-specific mean (group imputation)
iris_group_mean <- iris_missing %>%
  group_by(Species) %>%
  mutate(
    SepalLengthCm = ifelse(is.na(SepalLengthCm), 
                          mean(SepalLengthCm, na.rm = TRUE), 
                          SepalLengthCm),
    PetalWidthCm = ifelse(is.na(PetalWidthCm), 
                         mean(PetalWidthCm, na.rm = TRUE), 
                         PetalWidthCm)
  ) %>%
  ungroup()

miss_var_summary(iris_group_mean)
```

### Handling NAs During Analysis

```{r}
# Without na.rm - returns NA
mean(iris_missing$SepalLengthCm)

# With na.rm = TRUE - ignores NAs
mean(iris_missing$SepalLengthCm, na.rm = TRUE)

# Summary with NA handling
iris_missing %>%
  summarize(
    mean_sepal = mean(SepalLengthCm, na.rm = TRUE),
    median_sepal = median(SepalLengthCm, na.rm = TRUE),
    count_non_na = sum(!is.na(SepalLengthCm)),
    count_na = sum(is.na(SepalLengthCm))
  )
```

------------------------------------------------------------------------

## Part 4: Practice and Q&A (20 mins)

### Practice Tasks (based on iris data)

#### Task 1: Import and Explore
-   Import the iris dataset
-   Display the first 10 rows
-   Check the number of rows and columns

```{r}
# Solution
iris_practice <- read.csv("../data/Iris.csv")
head(iris_practice, 10)
dim(iris_practice)
```

#### Task 2: Filter and Select
-   Select only Iris-virginica flowers with PetalLengthCm > 6
-   Show only Species, SepalLengthCm, and PetalLengthCm columns

```{r}
# Solution
iris_practice %>%
  filter(Species == "Iris-virginica", PetalLengthCm > 6) %>%
  select(Species, SepalLengthCm, PetalLengthCm)
```

#### Task 3: Create New Variables
-   Create a variable for PetalRatio = PetalLengthCm / PetalWidthCm
-   Create a variable for SepalRatio = SepalLengthCm / SepalWidthCm

```{r}
# Solution
iris_practice <- iris_practice %>%
  mutate(
    PetalRatio = PetalLengthCm / PetalWidthCm,
    SepalRatio = SepalLengthCm / SepalWidthCm
  )

head(iris_practice)
```

#### Task 4: Group and Summarize
-   Group by Species
-   Calculate average SepalLengthCm, PetalLengthCm, and count per species

```{r}
# Solution
iris_practice %>%
  group_by(Species) %>%
  summarize(
    avg_sepal_length = mean(SepalLengthCm, na.rm = TRUE),
    avg_petal_length = mean(PetalLengthCm, na.rm = TRUE),
    count = n()
  )
```

#### Task 5: Complex Pipeline
-   Filter for flowers with SepalLengthCm > 6
-   Create a new variable TotalLength = SepalLengthCm + PetalLengthCm
-   Group by Species
-   Calculate average TotalLength
-   Sort by average TotalLength descending

```{r}
# Solution
iris_practice %>%
  filter(SepalLengthCm > 6) %>%
  mutate(TotalLength = SepalLengthCm + PetalLengthCm) %>%
  group_by(Species) %>%
  summarize(
    avg_total_length = mean(TotalLength, na.rm = TRUE),
    count = n()
  ) %>%
  arrange(desc(avg_total_length))
```

------------------------------------------------------------------------

## 📚 Further Resources (for Data Import & Wrangling)

-   Tidyverse Documentation: https://www.tidyverse.org/packages/
-   Importing Data in R: https://r4ds.hadley.nz/data-import.html
-   Data Transformation Cheatsheet: https://posit.co/resources/cheatsheets/