---
title: "Statistical Tests with R: T-tests, ANOVA, and Correlation"
author: "Mohan Khanal"
date: "2025-08-05"
format:
  html:
    toc: true
    toc-location: left
    toc-title: "Contents"
    toc-depth: 2
  pdf:
    toc: true
    number-sections: true
---
# Statistical Tests with R (T-tests, ANOVA, and correlation)


## Load required libraries 
```{r}
library(dplyr)
library(readr)
library(ggplot2)
```

```{r}
## Import dataset ----
car_data <- read.csv("D:/DemoPractices/R for Data Science/data/car data.csv")
#car_data <- read_csv("D:/DemoPracticedata/car data.csv")

# Preview data
head(car_data)
```

## Part 1: Exploratory Data Analysis 
### Cross-Tabulation 
Cross-tabulation summarizes the relationship between two categorical variables, showing counts and percentages.
Example: How do Fuel_Type and Transmission relate?
```{r}
#row percentage
crosstab_rw <- modelsummary::datasummary_crosstab(
  Fuel_Type ~ Transmission, 
  statistic = ~N +1 +Percent("row"),
  data = car_data
)
crosstab_rw
```


```{r}
#column percentage 
crosstab_cl <- modelsummary::datasummary_crosstab(
  Fuel_Type ~ Transmission,
  statistic = ~ N + 1+Percent("col"),
  data = car_data
)
crosstab_cl
```

### Correlation 
Correlation measures the strength and direction of the relationship between two numeric variables.

Example: Is there a relationship between Selling_Price and Kms_Driven?

```{r}
correlation <- cor.test(car_data$Selling_Price, car_data$Kms_Driven)
correlation
```

Interpretation:
Correlation coefficient (r): Positive (both increase together), negative (one increases, other decreases), or near zero (no linear relationship).
p-value < 0.05 indicates a statistically significant relationship.

## Part 2: T-tests
T-tests compare the means of two groups to determine if there is a statistically significant difference.
Example: Is the average Selling_Price different between manual and automatic cars?
```{r}
t_test_result <- t.test(Selling_Price ~ Transmission, data = car_data)
t_test_result
```

Interpretation:
p-value < 0.05: Significant difference in means.
Confidence interval: Shows the range of the true difference in means.
p-value > 0.05: No strong evidence of a difference.

##Part 3: ANOVA
ANOVA compares means across three or more groups.
Example: Does average Selling_Price differ among Fuel_Type categories?
```{r}
anova_result <- aov(Selling_Price ~ Transmission, data = car_data)
summary(anova_result)
```

### Post-hoc Test
```{r}

TukeyHSD(anova_result)
```
Interpretation:

p-value < 0.05: At least one group mean differs.
Tukey’s test: Identifies which specific groups differ.


## Part 4: Visualizations

```{r}
# Boxplot for T-test variable
boxplot(Selling_Price ~ Transmission, data = car_data,
        main = "Selling Price by Transmission",
        xlab = "Transmission", ylab = "Selling Price",
        col = "lightblue")

# Boxplot for ANOVA variable
boxplot(Selling_Price ~ Fuel_Type, data = car_data,
        main = "Selling Price by Fuel Type",
        xlab = "Fuel Type", ylab = "Selling Price",
        col = "orange")

# Scatter plot for correlation
plot(car_data$Present_Price, car_data$Selling_Price,
     main = "Present Price vs Selling Price",
     xlab = "Present Price", ylab = "Selling Price",
     pch = 19, col = "darkgreen")
```
