---
title: "4 Statistical Tests with R"
author: "Mohan Khanal"
date: "2026-01-17"
format:
  html:
    toc: true
    toc-location: left
    toc-title: "Contents"
    toc-depth: 2
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(readr)
options(warn = -1)
```

# 📊 Chapter 4: Statistical Tests with R

## 🕐 Agenda

-   **Part 1 (30 mins)** – Exploratory Data Analysis (Cross-tabulation & Correlation)
-   **Part 2 (30 mins)** – T-Tests (Comparing Two Groups)
-   **Part 3 (30 mins)** – ANOVA (Comparing Multiple Groups)
-   **Part 4 (20 mins)** – Visualizations & Interpretation
-   **Part 5 (10 mins)** – Practice and Q&A

------------------------------------------------------------------------

## Setup: Load Data and Libraries

```{r, warning=FALSE, message=FALSE}
# Load required libraries
library(dplyr)
library(readr)
library(ggplot2)

# Import car sales dataset

car_data <- read.csv("~/Documents/khanalmohan.github.io/data/car_data.csv")

# Check structure
str(car_data)

```

### Understanding the Dataset

Our car sales dataset contains:
- **Selling_Price**: Price at which car was sold
- **Present_Price**: Current market price
- **Kms_Driven**: Total kilometers driven
- **Fuel_Type**: Petrol, Diesel, or CNG
- **Transmission**: Manual or Automatic
- **Owner**: Number of previous owners
- **Year**: Year of manufacture

------------------------------------------------------------------------

## Part 1: Exploratory Data Analysis (30 mins)

### Cross-Tabulation (Contingency Tables)

Cross-tabulation shows the relationship between two categorical variables.

```{r}
# Basic cross-tabulation with counts
table(car_data$Fuel_Type, car_data$Transmission)

# Add row and column totals
addmargins(table(car_data$Fuel_Type, car_data$Transmission))
```

#### Enhanced Cross-Tabulation with Percentages

```{r, warning=FALSE, message=FALSE}
library(modelsummary)

# Row percentages - "Of cars with this fuel type, what % have each transmission?"
crosstab_row <- datasummary_crosstab(
  Fuel_Type ~ Transmission, 
  statistic = ~ N + 1 + Percent("row"),
  data = car_data
)
crosstab_row
```

```{r}
# Column percentages - "Of cars with this transmission, what % have each fuel type?"
crosstab_col <- datasummary_crosstab(
  Fuel_Type ~ Transmission,
  statistic = ~ N + 1 + Percent("col"),
  data = car_data
)
crosstab_col
```

**💡 Interpretation Tips:**
- Look for cells with notably high or low counts
- Row % shows distribution within each row category
- Column % shows distribution within each column category

### Correlation Analysis

Correlation measures the strength of relationship between two numeric variables.

```{r}
# Correlation between Selling Price and Kilometers Driven
cor.test(car_data$Selling_Price, car_data$Kms_Driven)
```

**Understanding Correlation:**
- **Range**: -1 to +1
- **Positive (0 to +1)**: Variables increase together
- **Negative (-1 to 0)**: One increases, other decreases
- **Near zero**: No linear relationship

**Strength Guide:**
- 0.0 to 0.3: Weak
- 0.3 to 0.7: Moderate
- 0.7 to 1.0: Strong

```{r}
# Correlation matrix for multiple variables
numeric_vars <- car_data %>% 
  select(Selling_Price, Present_Price, Kms_Driven, Year)

cor_matrix <- cor(numeric_vars, use = "complete.obs")
round(cor_matrix, 3)
```

### Visualize Correlation

```{r}
# Scatter plot with correlation
ggplot(car_data, aes(x = Present_Price, y = Selling_Price)) +
  geom_point(alpha = 0.5, color = "darkgreen") +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(
    title = "Present Price vs Selling Price",
    subtitle = paste("Correlation:", round(cor(car_data$Present_Price, car_data$Selling_Price), 3)),
    x = "Present Price (in lakhs)",
    y = "Selling Price (in lakhs)"
  ) +
  theme_minimal()
```

------------------------------------------------------------------------

## Part 2: T-Tests (30 mins)

### What is a T-Test?

T-tests compare the **means** of two groups to see if they're significantly different.

**Example Question:** Do automatic cars have different selling prices than manual cars?

### Hypotheses

- **Null Hypothesis (H₀)**: No difference in average selling prices
- **Alternative Hypothesis (H₁)**: There IS a difference in average selling prices

### Running the T-Test

```{r}
# Independent samples t-test
t_test_result <- t.test(Selling_Price ~ Transmission, data = car_data)
t_test_result
```

### Interpreting T-Test Results

**Key Components:**
1. **t-statistic**: How many standard errors separate the two means
2. **p-value**: 
   - p < 0.05 → **Significant difference** (reject null hypothesis)
   - p > 0.05 → No significant difference
3. **Confidence interval**: Range where true difference likely falls
4. **Sample means**: Average for each group

### Visualizing T-Test Results

```{r}
# Boxplot comparison
ggplot(car_data, aes(x = Transmission, y = Selling_Price, fill = Transmission)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Selling Price by Transmission Type",
    subtitle = paste("p-value:", round(t_test_result$p.value, 4)),
    x = "Transmission Type",
    y = "Selling Price (in lakhs)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

### Calculate Descriptive Stats by Group

```{r}
car_data %>%
  group_by(Transmission) %>%
  summarize(
    Mean = mean(Selling_Price),
    SD = sd(Selling_Price),
    Median = median(Selling_Price),
    Count = n()
  )
```

### Check T-Test Assumptions

```{r}
# Visual check: Q-Q plots for normality
par(mfrow = c(1, 2))
qqnorm(car_data$Selling_Price[car_data$Transmission == "Manual"], 
       main = "Q-Q Plot: Manual")
qqline(car_data$Selling_Price[car_data$Transmission == "Manual"])

qqnorm(car_data$Selling_Price[car_data$Transmission == "Automatic"], 
       main = "Q-Q Plot: Automatic")
qqline(car_data$Selling_Price[car_data$Transmission == "Automatic"])
```

------------------------------------------------------------------------

## Part 3: ANOVA (30 mins)

### What is ANOVA?

ANOVA (Analysis of Variance) compares means across **three or more groups**.

**Example Question:** Do selling prices differ among different fuel types (Petrol, Diesel, CNG)?

### Hypotheses

- **Null Hypothesis (H₀)**: All group means are equal
- **Alternative Hypothesis (H₁)**: At least one group mean differs

### Running ANOVA

```{r}
# One-way ANOVA
anova_result <- aov(Selling_Price ~ Fuel_Type, data = car_data)
summary(anova_result)
```

### Interpreting ANOVA Results

**Key Components:**
1. **F-statistic**: Ratio of between-group to within-group variance
2. **p-value (Pr(>F))**: 
   - p < 0.05 → At least one group differs significantly
   - p > 0.05 → No significant differences detected

### Post-Hoc Tests (Which groups differ?)

If ANOVA is significant, use **Tukey's HSD** to find which specific groups differ.

```{r}
# Tukey's Honestly Significant Difference
tukey_result <- TukeyHSD(anova_result)
tukey_result
```

**💡 Reading Tukey Results:**
- **diff**: Difference in means between groups
- **p adj**: Adjusted p-value
  - p adj < 0.05 → Groups significantly differ
  - p adj > 0.05 → No significant difference
- If confidence interval includes 0 → not significant

### Visualize Tukey Results

```{r}
# Plot Tukey confidence intervals
plot(tukey_result, las = 1)
```

### Group Statistics

```{r}
car_data %>%
  group_by(Fuel_Type) %>%
  summarize(
    Mean = mean(Selling_Price),
    SD = sd(Selling_Price),
    Median = median(Selling_Price),
    Count = n()
  )
```

------------------------------------------------------------------------

## Part 4: Visualizations & Interpretation (20 mins)

### Comprehensive Comparison Plots

```{r}
# Create three-panel comparison
par(mfrow = c(1, 3))

# 1. T-test visualization
boxplot(Selling_Price ~ Transmission, data = car_data,
        main = "T-Test: Transmission",
        xlab = "Transmission", 
        ylab = "Selling Price",
        col = c("lightblue", "lightcoral"))

# 2. ANOVA visualization
boxplot(Selling_Price ~ Fuel_Type, data = car_data,
        main = "ANOVA: Fuel Type",
        xlab = "Fuel Type", 
        ylab = "Selling Price",
        col = c("orange", "lightgreen", "yellow"))

# 3. Correlation visualization
plot(car_data$Present_Price, car_data$Selling_Price,
     main = "Correlation Analysis",
     xlab = "Present Price", 
     ylab = "Selling Price",
     pch = 19, col = "darkgreen")
abline(lm(Selling_Price ~ Present_Price, data = car_data), 
       col = "red", lwd = 2)
```

### Enhanced ANOVA Visualization

```{r}
# Boxplot with individual points
ggplot(car_data, aes(x = Fuel_Type, y = Selling_Price, fill = Fuel_Type)) +
  geom_boxplot(alpha = 0.7, outlier.color = "red") +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
  stat_summary(fun = mean, geom = "point", 
               shape = 23, size = 3, fill = "white", color = "black") +
  labs(
    title = "Selling Price by Fuel Type",
    subtitle = "Diamonds show group means",
    x = "Fuel Type",
    y = "Selling Price (in lakhs)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

### How to Report Results

**T-Test Example:**
"Automatic transmission cars (M = 7.5 lakhs) had significantly higher selling prices than manual transmission cars (M = 4.3 lakhs), t(298) = 6.82, p < .001."

**ANOVA Example:**
"Selling prices differed significantly across fuel types, F(2, 297) = 12.45, p < .001. Post-hoc tests showed diesel cars were more expensive than petrol (p = .002) and CNG cars (p < .001)."

**Correlation Example:**
"There was a strong positive correlation between present price and selling price, r = .89, p < .001."

------------------------------------------------------------------------

## Part 5: Practice and Q&A (10 mins)

### Practice Tasks

#### Task 1: Cross-Tabulation
Create a cross-tabulation of `Owner` and `Fuel_Type`. What patterns do you see?

```{r}
# Solution
table(car_data$Owner, car_data$Fuel_Type)
```

#### Task 2: Correlation
Calculate the correlation between `Year` and `Selling_Price`. Interpret the result.

```{r}
# Solution
cor.test(car_data$Year, car_data$Selling_Price)

# Visualization
ggplot(car_data, aes(x = Year, y = Selling_Price)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Year vs Selling Price") +
  theme_minimal()
```

#### Task 3: T-Test
Test if there's a difference in kilometers driven between Petrol and Diesel cars.

```{r}
# Solution
# Filter for just Petrol and Diesel
petrol_diesel <- car_data %>% 
  filter(Fuel_Type %in% c("Petrol", "Diesel"))

t.test(Kms_Driven ~ Fuel_Type, data = petrol_diesel)
```

#### Task 4: ANOVA
Test if `Present_Price` differs across different `Owner` categories.

```{r}
# Solution
anova_owner <- aov(Present_Price ~ factor(Owner), data = car_data)
summary(anova_owner)

# If significant, run Tukey
TukeyHSD(anova_owner)
```

#### Task 5: Complete Analysis
Compare `Kms_Driven` between Manual and Automatic transmission:
1. Calculate descriptive statistics
2. Run t-test
3. Create visualization

```{r}
# Solution

# 1. Descriptive statistics
car_data %>%
  group_by(Transmission) %>%
  summarize(
    Mean_Kms = mean(Kms_Driven),
    SD_Kms = sd(Kms_Driven),
    Median_Kms = median(Kms_Driven),
    N = n()
  )

# 2. T-test
t_test_kms <- t.test(Kms_Driven ~ Transmission, data = car_data)
t_test_kms

# 3. Visualization
ggplot(car_data, aes(x = Transmission, y = Kms_Driven, fill = Transmission)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Kilometers Driven by Transmission Type",
    subtitle = paste("p-value:", round(t_test_kms$p.value, 4)),
    x = "Transmission",
    y = "Kilometers Driven"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

### Challenge: Multi-Step Analysis

Investigate the relationship between car age and selling price:
1. Create an `Age` variable (2026 - Year)
2. Calculate correlation with `Selling_Price`
3. Create scatter plot with trend line
4. Group cars into "New" (Age < 5) and "Old" (Age >= 5)
5. Run t-test comparing selling prices

```{r}
# Solution

# 1. Create Age variable
car_data <- car_data %>%
  mutate(Age = 2026 - Year)

# 2. Correlation
cor_test <- cor.test(car_data$Age, car_data$Selling_Price)
cor_test

# 3. Scatter plot
ggplot(car_data, aes(x = Age, y = Selling_Price)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(
    title = "Car Age vs Selling Price",
    subtitle = paste("Correlation:", round(cor_test$estimate, 3)),
    x = "Car Age (years)",
    y = "Selling Price (in lakhs)"
  ) +
  theme_minimal()

# 4. Create age groups
car_data <- car_data %>%
  mutate(Age_Group = ifelse(Age < 5, "New", "Old"))

# 5. T-test
t.test(Selling_Price ~ Transmission, data = car_data)

# Visualization
ggplot(car_data, aes(x = Age_Group, y = Selling_Price, fill = Age_Group)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Selling Price: New vs Old Cars",
    x = "Age Group",
    y = "Selling Price (in lakhs)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

------------------------------------------------------------------------

## 📚 Key Takeaways

✅ **Cross-tabulation** shows relationships between categorical variables  
✅ **Correlation** measures strength of relationship between numeric variables (-1 to +1)  
✅ **T-tests** compare means of two groups (p < 0.05 = significant)  
✅ **ANOVA** compares means of 3+ groups  
✅ **Tukey's HSD** identifies which specific groups differ after ANOVA  
✅ Always visualize results with boxplots and scatter plots  
✅ Check assumptions: normality, equal variance  
✅ Report: test statistic, p-value, means, and interpretation  

------------------------------------------------------------------------

## 🎯 Statistical Decision Guide

| **Question** | **Test to Use** |
|-------------|----------------|
| Compare 2 groups (numeric outcome) | T-test |
| Compare 3+ groups (numeric outcome) | ANOVA + Tukey |
| Relationship between 2 numeric variables | Correlation |
| Relationship between 2 categorical variables | Cross-tabulation / Chi-square |

------------------------------------------------------------------------

## 📖 Resources

-   Quick-R Statistical Tests: https://www.statmethods.net/stats/
-   R Statistics Documentation: `?t.test`, `?aov`, `?cor.test`
-   ANOVA Tutorial: https://www.datanovia.com/en/lessons/anova-in-r/
-   Correlation Guide: https://www.statology.org/correlation-in-r/

------------------------------------------------------------------------

## 🔜 Next Chapter Preview

**Chapter 5: Regression Analysis**
- Simple and multiple linear regression
- Model building and interpretation
- Prediction and forecasting
- Model diagnostics
