---
title: "1 Introduction to R and RStudio"
subtitle: "January 05-06, 2026 | 7:00 - 8:30 PM"
author: "Mohan Khanal"
---

# Welcome to R Programming! ðŸŽ‰

## What is R?

R is a powerful programming language and environment for:

- Statistical computing and analysis
- Data visualization
- Data manipulation and cleaning
- Machine learning and predictive modeling
- Reproducible research

## Why Learn R?

- **Free and Open Source**: No licensing costs
- **Extensive Packages**: Over 19,000 packages available
- **Strong Community**: Active support and resources
- **Industry Standard**: Used in academia, healthcare, finance, tech
- **Data Science Ready**: Built specifically for data analysis

---

# Getting Started with RStudio

## RStudio Interface Overview

RStudio has 4 main panes:

1. **Source/Editor Pane** (Top Left): Where you write and save your code
2. **Console Pane** (Bottom Left): Where code executes and shows output
3. **Environment/History Pane** (Top Right): Shows variables and command history
4. **Files/Plots/Packages/Help Pane** (Bottom Right): Multiple tabs for different functions

### Quick Tips:
- Press **Ctrl + Enter** (Windows) or **Cmd + Enter** (Mac) to run current line
- Press **Ctrl + Shift + Enter** to run entire script
- Use **#** for comments (code that won't execute)

---

# Your First R Code

## The Classic "Hello World"

```{r}
print("Hello World")
```

## Built-in Constants

```{r}
# Pi is a built-in constant in R
pi

# R has many mathematical constants
exp(1)  # Euler's number (e)
```

---

# Basic Math Operations

## Arithmetic Operators

```{r}
# Addition
2 + 2

# Subtraction
10 - 3

# Multiplication
4 * 5

# Division
20 / 4

# Exponentiation (Power)
2^3

# Modulo (Remainder)
10 %% 3

# Integer Division
10 %/% 3
```

## Order of Operations (PEMDAS)

```{r}
# R follows standard mathematical order
2 + 3 * 4      # Multiplication first

# Use parentheses to change order
(2 + 3) * 4    # Addition first
```

---

# Data Types in R

## 1. Numeric (Double)

The default for numbers with decimals:

```{r}
x <- 5
x
class(x)
typeof(x)

# Decimal numbers
m <- 0.03
class(m)

# Scientific notation
large_num <- 1.5e6  # 1,500,000
large_num
```

## 2. Integer

Whole numbers (add 'L' suffix):

```{r}
y <- 444L
class(y)
typeof(y)

# Check if it's an integer
is.integer(y)
is.integer(5)      # FALSE - default is numeric
is.integer(5L)     # TRUE
```

## 3. Character (String)

Text data enclosed in quotes:

```{r}
z <- "This is our first data camp"
z
class(z)

# Single or double quotes both work
name <- 'John'
greeting <- "Hello, World!"

# Combine strings
paste("Hello", "R", "Programming")
paste0("Hello", "World")  # No space
```

## 4. Logical (Boolean)

TRUE or FALSE values:

```{r}
aa <- FALSE
class(aa)

bb <- TRUE
bb

# Logical operations
TRUE & FALSE   # AND
TRUE | FALSE   # OR
!TRUE          # NOT

# Comparisons return logical values
5 > 3
10 == 10
7 != 8
```

## Quick Type Check Summary

```{r}
# Check multiple types at once
check_value <- 42

is.numeric(check_value)
is.character(check_value)
is.logical(check_value)
```

---

# Variables and Assignment

## Creating Variables

```{r}
# Use <- or = for assignment (â† is preferred in R)
k <- 11
l <- 9

# View variables
k
l

# Use variables in calculations
k + l
k * l
k / l

# Update variables
k <- k + 5
k
```

## Variable Naming Rules

```{r}
# Good variable names
student_age <- 20
studentAge <- 20      # camelCase
student.age <- 20     # dot notation

# Names can include numbers (but not start with them)
data1 <- 100
# 1data <- 100  # This would cause an ERROR

# Case sensitive
Value <- 10
value <- 20
Value == value  # FALSE
```

## Best Practices

- Use descriptive names: `student_score` instead of `x`
- Be consistent with naming style
- Avoid special characters except `.` and `_`
- Don't overwrite built-in functions (like `mean`, `sum`, `c`)

---

# Using Functions

## What are Functions?

Functions are pre-built commands that perform specific tasks.

Syntax: `function_name(arguments)`

## Basic Mathematical Functions

```{r}
# Square root
sqrt(16)

# Absolute value
abs(-5)

# Rounding
round(3.14159, 2)  # Round to 2 decimal places

# Logarithms
log(10)      # Natural log
log10(100)   # Base 10 log

# Exponential
exp(2)
```

## The c() Function (Concatenate)

Creates vectors (collections of values):

```{r}
# Create a vector of numbers
numbers <- c(4, 5, 6)
numbers

# Calculate mean
mean(c(4, 5, 6))

# Calculate sum
sum(c(4, 5, 6, 7))

# Create sequences
sum(1:10)      # Sum of 1 to 10
mean(1:10)     # Mean of 1 to 10

# More with vectors
ages <- c(23, 45, 67, 34, 29)
min(ages)
max(ages)
median(ages)
length(ages)   # How many elements
```

## Statistical Functions

```{r}
# Create sample data
scores <- c(85, 90, 78, 92, 88, 76, 95)

# Descriptive statistics
mean(scores)       # Average
median(scores)     # Middle value
sd(scores)         # Standard deviation
var(scores)        # Variance
range(scores)      # Min and max
summary(scores)    # Quick summary
```

---

# Sequences and Patterns

## Creating Sequences

```{r}
# Using colon operator
1:10
5:15
10:1  # Descending

# Using seq() function
seq(1, 10)                    # Same as 1:10
seq(0, 100, by = 10)          # Specify step
seq(0, 1, by = 0.1)           # Decimal steps
seq(0, 1, length.out = 11)    # Specify length

# Repeating values
rep(5, times = 3)             # Repeat 5 three times
rep(1:3, times = 3)           # Repeat sequence
rep(1:3, each = 3)            # Repeat each element
```

---

# Getting Help in R

## Help Functions

```{r eval=FALSE}
# Get help on a function
?mean
help(sum)

# Search for topics
??regression
help.search("linear model")

# Examples of function usage
example(mean)

# See arguments of a function
args(seq)
```

## Useful Commands

```{r}
# List objects in environment
ls()

# Remove an object
# rm(x)

# Clear entire environment
# rm(list = ls())

# Check current working directory
getwd()

# Set working directory (example - modify path)
# setwd("C:/Users/YourName/Documents/R_Training")
```

---

# Practice Exercises

## Exercise 1: Basic Calculations

Calculate the following and store in variables:

1. The sum of 15 and 27
2. 8 raised to the power of 3
3. The square root of 144

```{r}
# Your code here
sum_result <- 15 + 27
power_result <- 8^3
sqrt_result <- sqrt(144)

# View results
sum_result
power_result
sqrt_result
```

## Exercise 2: Working with Vectors

Create a vector of temperatures: 72, 68, 75, 80, 77

Calculate:
- Mean temperature
- Highest temperature
- Temperature range

```{r}
# Your code here
temperatures <- c(72, 68, 75, 80, 77)

mean(temperatures)
max(temperatures)
range(temperatures)
```

## Exercise 3: Data Types

Identify the data type of the following:

```{r}
var1 <- 42
var2 <- "Data Science"
var3 <- TRUE
var4 <- 3.14159

class(var1)
class(var2)
class(var3)
class(var4)
```

---

# Key Takeaways ðŸ“

1. **R is a powerful tool** for data analysis and statistics
2. **RStudio** provides an excellent interface for R programming
3. **Data types** matter: numeric, integer, character, logical
4. **Variables** store values using `<-` operator
5. **Functions** perform operations: `mean()`, `sum()`, `c()`, etc.
6. **Vectors** are collections of values created with `c()`
7. **Help is available** using `?function_name`

---

# Homework / Practice

Before our next session, practice:

1. Install R and RStudio on your computer (if not already done)
2. Create 5 different variables with different data types
3. Perform at least 10 different mathematical operations
4. Create 3 vectors and calculate their mean, median, and standard deviation
5. Experiment with the `seq()` and `rep()` functions
6. Use the `?` operator to explore 3 new functions

---

# Next Session Preview

**January 07, 2026: Data Import and Data Wrangling**

We'll cover:
- Reading CSV, Excel, and other data files
- Introduction to the `dplyr` package
- Data cleaning and manipulation techniques
- Handling missing data (NA values)

**Prepare:** Make sure you have a CSV file ready to import (any data will work!)

---

# Resources

- [R Documentation](https://www.r-project.org/)
- [RStudio Cheatsheets](https://posit.co/resources/cheatsheets/)
- [R for Data Science (Free Book)](https://r4ds.had.co.nz/)
- [Stack Overflow - R Tag](https://stackoverflow.com/questions/tagged/r)

---

**Questions? Feel free to ask during the session or practice at home!** ðŸ’ª

*Happy Coding!* ðŸš€