---
title: "Digital Divide: Facebook Penetration and Socioeconomic Development"
subtitle: "An Analysis Using Facebook Marketing API and World Bank Data"
author: "Mohan Khanal"
date: last-modified
date-format: "MMMM D, YYYY"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    theme: cosmo
    number-sections: true
    smooth-scroll: true
    self-contained: false
execute:
  warning: false
  message: false
  cache: false
bibliography: references.bib
---

# Introduction

This analysis explores the relationship between Facebook penetration and socioeconomic indicators across countries using data from Facebook's Marketing API and the World Bank's World Development Indicators (WDI). 

The study examines:

- Facebook user distribution across regions
- Gender disparities in Facebook usage
- Correlation between Facebook penetration and internet connectivity
- Relationship between social media adoption and economic development

## Data Sources

**Facebook Marketing API**: Provides estimates of monthly active users (MAU) by location and demographic characteristics. We use the `rsocialwatcher` R package to query this data.

**World Development Indicators**: The World Bank's comprehensive database of development indicators, accessed through the `WDI` R package.

# Setup

## Load Required Packages

```{r}
#| label: setup
#| code-fold: false

# Install packages if needed
# install.packages(c("rsocialwatcher", "WDI", "dplyr", "tidyr", "ggplot2", 
#                    "janitor", "ggpubr", "knitr", "kableExtra", "sf", "rnaturalearth"))

library(rsocialwatcher)
library(WDI)
library(dplyr)
library(tidyr)
library(ggplot2)
library(janitor)
library(ggpubr)
library(knitr)
library(kableExtra)
```

## Facebook API Credentials

::: {.callout-note}
To replicate this analysis, you'll need:

1. A Facebook Developer account
2. A Facebook Ad Account
3. Access token from Facebook Graph API
4. Ad Account ID (Creation Act)

Visit [Facebook for Developers](https://developers.facebook.com/) to set up your credentials.
:::

```{r}
#| label: credentials
#| eval: false
#| code-fold: false

# Replace with your actual credentials
TOKEN        <- "YOUR-TOKEN-HERE"
CREATION_ACT <- "YOUR-CREATION-ACT-HERE"
VERSION      <- "v19.0"  # Check latest API version
```

```{r}
#| label: credentials-real
#| include: false

# Actual credentials (not shown in output)
TOKEN        <- "EAANhP2mkZCosBO05ZAT7ZBqgPFuwR4JZAvc4ZBwaZAUHGvVDhZBSLWJ3C0ycP5oplTOADnkBPqradiTPGmybRYITWAyIdI2UD0FCUhqHn75BukzjT2ViEa7zIzCk9z5cUGEApyRUZCfKuplcHwh5bjj5nPEsDyC8uzjpwJrdbJWjQP1T2jpKdAxZAxTVcPCCzI4JbkfMxKEfk"
CREATION_ACT <- "843351257616092"
VERSION      <- "v19.0"
```

# Data Collection

## World Bank Development Indicators

We collect key socioeconomic indicators:

- `SP.POP.TOTL`: Total population
- `SP.POP.TOTL.MA.ZS`: Male population (% of total)
- `SP.POP.TOTL.FE.ZS`: Female population (% of total)
- `NY.GDP.PCAP.CD`: GDP per capita (current US$)
- `IT.NET.USER.ZS`: Individuals using the Internet (% of population)

```{r}
#| label: wdi-data
#| eval: false

# Query World Bank data
wdi_df <- WDI(
  country = "all",
  indicator = c(
    "SP.POP.TOTL",
    "SP.POP.TOTL.MA.ZS",
    "SP.POP.TOTL.FE.ZS",
    "NY.GDP.PCAP.CD",
    "IT.NET.USER.ZS"
  ),
  start = 2021,
  end = 2021,
  extra = TRUE
)

# Focus on a specific region (change as needed)
wdi_df <- wdi_df %>%
  filter(region == "South Asia")
```

## Facebook Marketing API Data

```{r}
#| label: fb-data
#| eval: false

# Query Facebook data
fb_df <- query_fb_marketing_api(
  location_unit_type = "countries",
  location_keys      = map_param_vec(wdi_df$iso2c),
  gender             = map_param(1, 2),  # 1 = male, 2 = female
  version            = VERSION,
  creation_act       = CREATION_ACT,
  token              = TOKEN
)
```

::: {.callout-tip}
## API Rate Limits

The Facebook Marketing API has rate limits. For large queries:

- Use batching with `map_param_vec()`
- Add delays between requests if needed
- Cache results locally to avoid repeated queries
:::

# Data Processing

## Merge and Clean Data

```{r}
#| label: data-cleaning
#| eval: false

fb_clean_df <- fb_df %>%
  # Rename for joining
  rename(iso2c = location_keys) %>%
  
  # Recode gender variable
  mutate(gender = case_when(
    gender == "1" ~ "fb_male",
    gender == "2" ~ "fb_female"
  )) %>%
  
  # Pivot to wide format
  pivot_wider(
    id_cols = iso2c,
    names_from = gender,
    values_from = estimate_mau_upper_bound
  ) %>%
  
  # Join with World Bank data
  left_join(wdi_df, by = "iso2c") %>%
  
  # Clean column names
  clean_names() %>%
  
  # Calculate derived variables
  mutate(
    fb_total = fb_female + fb_male,
    fb_per_female = (fb_female / fb_total) * 100,
    wdi_per_female = sp_pop_totl_fe_zs,
    per_fb_pop = (fb_total / sp_pop_totl) * 100,
    fb_penetration_internet = (fb_total / (sp_pop_totl * it_net_user_zs / 100)) * 100
  )
```

## Summary Statistics

```{r}
#| label: summary-table
#| eval: false

fb_clean_df %>%
  select(
    country,
    fb_total,
    per_fb_pop,
    fb_per_female,
    it_net_user_zs,
    ny_gdp_pcap_cd
  ) %>%
  arrange(desc(per_fb_pop)) %>%
  kable(
    digits = 2,
    col.names = c(
      "Country",
      "FB Users (Total)",
      "FB Penetration (%)",
      "Female Users (%)",
      "Internet Users (%)",
      "GDP per Capita ($)"
    ),
    caption = "Facebook Penetration and Socioeconomic Indicators"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  scroll_box(height = "400px")
```

# Analysis and Visualization

## Theme Setup

```{r}
#| label: plot-theme
#| code-fold: false

# Custom theme for plots
p_theme <- theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(face = "italic", size = 11, color = "grey40"),
    axis.text = element_text(color = "black", size = 10),
    axis.title = element_text(size = 11, face = "bold"),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "grey80", fill = NA, linewidth = 0.5)
  )
```

## Facebook Penetration Distribution

```{r}
#| label: fig-distribution
#| eval: false
#| fig-cap: "Distribution of Facebook penetration across countries"
#| fig-width: 10
#| fig-height: 4

p_1a <- fb_clean_df %>%
  ggplot(aes(x = per_fb_pop)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 15,
    fill = "#4267B2",
    color = "white",
    alpha = 0.8
  ) +
  geom_density(color = "#1a3a52", linewidth = 1, alpha = 0.3) +
  labs(
    x = "Facebook Penetration (% of population)",
    y = "Density",
    title = "Distribution of Facebook Penetration"
  ) +
  p_theme

p_1a
```

## Facebook vs Internet Connectivity

```{r}
#| label: fig-internet
#| eval: false
#| fig-cap: "Relationship between internet connectivity and Facebook penetration"
#| fig-width: 8
#| fig-height: 6

p_1b <- fb_clean_df %>%
  ggplot(aes(x = it_net_user_zs, y = per_fb_pop)) +
  geom_point(
    aes(size = sp_pop_totl / 1e6),
    fill = "#4267B2",
    color = "white",
    pch = 21,
    alpha = 0.7
  ) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    color = "#1a3a52",
    fill = "#4267B2",
    alpha = 0.2
  ) +
  labs(
    x = "Internet Users (% of population)",
    y = "Facebook Penetration (% of population)",
    title = "Internet Connectivity vs Facebook Penetration",
    subtitle = "Point size represents total population (millions)",
    size = "Population (M)"
  ) +
  p_theme

p_1b
```

## Facebook vs Economic Development

```{r}
#| label: fig-gdp
#| eval: false
#| fig-cap: "Relationship between GDP per capita and Facebook penetration"
#| fig-width: 8
#| fig-height: 6

p_1c <- fb_clean_df %>%
  ggplot(aes(x = ny_gdp_pcap_cd, y = per_fb_pop)) +
  geom_point(
    aes(size = sp_pop_totl / 1e6),
    fill = "#4267B2",
    color = "white",
    pch = 21,
    alpha = 0.7
  ) +
  geom_smooth(
    method = "loess",
    se = TRUE,
    color = "#1a3a52",
    fill = "#4267B2",
    alpha = 0.2
  ) +
  scale_x_log10(labels = scales::dollar_format()) +
  labs(
    x = "GDP per Capita (log scale, US$)",
    y = "Facebook Penetration (% of population)",
    title = "Economic Development vs Facebook Penetration",
    subtitle = "Point size represents total population (millions)",
    size = "Population (M)"
  ) +
  p_theme

p_1c
```

## Gender Disparities

```{r}
#| label: fig-gender
#| eval: false
#| fig-cap: "Gender parity in Facebook usage vs population"
#| fig-width: 8
#| fig-height: 6

p_gender <- fb_clean_df %>%
  ggplot(aes(x = wdi_per_female, y = fb_per_female)) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed",
    color = "grey50",
    linewidth = 1
  ) +
  geom_point(
    aes(size = fb_total / 1e6),
    fill = "#4267B2",
    color = "white",
    pch = 21,
    alpha = 0.7
  ) +
  labs(
    x = "Female Population (% of total)",
    y = "Female Facebook Users (% of total FB users)",
    title = "Gender Parity: Population vs Facebook Usage",
    subtitle = "Dashed line represents perfect parity. Point size = total FB users (millions)",
    size = "FB Users (M)"
  ) +
  coord_equal(xlim = c(45, 55), ylim = c(25, 55)) +
  p_theme

p_gender
```

## Combined Dashboard

```{r}
#| label: fig-dashboard
#| eval: false
#| fig-cap: "Comprehensive dashboard of Facebook penetration metrics"
#| fig-width: 12
#| fig-height: 8

ggarrange(
  p_1a, p_1b, p_1c, p_gender,
  ncol = 2,
  nrow = 2,
  labels = c("A", "B", "C", "D"),
  common.legend = FALSE
)
```

# Statistical Analysis

## Correlation Analysis

```{r}
#| label: correlation
#| eval: false

# Select numeric variables for correlation
cor_vars <- fb_clean_df %>%
  select(
    per_fb_pop,
    it_net_user_zs,
    ny_gdp_pcap_cd,
    fb_per_female,
    wdi_per_female
  ) %>%
  na.omit()

# Calculate correlation matrix
cor_matrix <- cor(cor_vars, use = "complete.obs")

# Display
kable(
  cor_matrix,
  digits = 3,
  caption = "Correlation Matrix"
) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Regression Analysis

```{r}
#| label: regression
#| eval: false

# Simple linear model
model1 <- lm(per_fb_pop ~ it_net_user_zs + log(ny_gdp_pcap_cd), 
             data = fb_clean_df)

# Display results
summary(model1)
```

# Discussion

## Key Findings

1. **Digital Divide**: Facebook penetration varies significantly across countries, closely tracking internet connectivity rates.

2. **Economic Correlation**: Higher GDP per capita is associated with greater Facebook adoption, though the relationship is not strictly linear.

3. **Gender Gap**: In many countries, female representation on Facebook differs from their proportion in the general population.

4. **Infrastructure Matters**: Internet infrastructure appears to be a primary determinant of social media adoption.

## Limitations

- **API Data**: Facebook Marketing API provides estimates, not exact counts
- **Temporal**: Cross-sectional analysis; longitudinal trends not captured
- **Selection Bias**: Only includes Facebook; other platforms not considered
- **Causality**: Correlational analysis; cannot infer causal relationships

## Future Directions

- Extend analysis to other regions (Sub-Saharan Africa, Latin America)
- Include additional social media platforms
- Examine behavioral indicators (interests, purchasing power)
- Time-series analysis of adoption patterns
- Incorporate policy and regulatory variables

# Reproducibility

## Session Information

```{r}
#| label: session-info

sessionInfo()
```

## Data Availability

- **Facebook data**: Requires API credentials
- **World Bank data**: Publicly available via WDI package
- **Analysis code**: Available at [github.com/khanalmohan/replication-studies](https://github.com/khanalmohan)

# References

::: {#refs}
:::

---

**Last updated**: `r Sys.Date()`

**License**: CC BY 4.0
